Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-24 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_331cf}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Technically, the introduction of this additional random variable doubles
the space of the hidden states and therefore quadruples the size of the
transition matrix \(\mathbf{A}\). However, since we assume no
correlation between the tumor's lateralization and the metastases in the
LNLs, we can evolve the two parts separately.

The probabilities to find a patient with a clearly lateralized tumor or
one that extends over the mid-sagittal line after \(t\) time-steps are
then given by

@@

To get the joint probability over the ipsi- and contralateral hidden
states, as well as the state of the tumor's midline extension
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
we simply add the above terms to the marginalization in
equation~\ref{eq-bilateral-marginal}:

\[
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\]

Again, this can be written algebraically, by defining the vector
\(P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})\),
to achieve the same form as in
equation~\ref{eq-bilateral-marginal-algebra}:

\[
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
\]

@@

\caption{}\label{T_d4a82}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

\caption{\label{fig-model-prevalences-with-ipsi}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences. Shown are four
scenarios, all including contralateral LNL II involvement: Early
T-category and an ipsilateral N0 neck (green), early T-category and
ipsilateral LNL II involvement (blue), as well as the same two scenarios
but for advanced T-category (orange and red). The figure shows that the
model is capable of describing the correlation between ipsi- and
contralateral involvement. Although for the scenario of LNL II
involvement in both sides, the prediction's split between early and
advanced T-category is not large enough.}

@@

In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
ability to capture the correlation between ipsi- and contralateral
involvement. It shows that the prevalence of metastases in the two sides
of the neck is correlated via the time of diagnosis, despite the model
not having any direct connections between the two side. However, there
are some small discrepancies in the model's prediction: When considering
the scenario of LNL II involvement in the ipsi- \emph{and} contralateral
side, it cannot quite capture the split between early and late
T-category. The model slightly overestimates the prevalence of this
scenario for early T-category patients and underestimates it for
advanced T-category.

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-risks.png}

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult contralateral LNL II invovlement, shown for some combinations of
T-category, tumor lateralization, and ipsilateral clinical involvement.
The contralateral side was always assumed to be clinically negative.}

@@

Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
in contralateral LNL II for four interesting combinations of these three
risk factors. This risk is computed \emph{given} a CT diagnosis
assessing the per-LNL clinical involvement for which we assumed a
sensitivity of 81\% and a specificity of 76\%
\citep{de_bondt_detection_2007}.

The variable impacting the prediction for contralateral involvement in
our model is the tumor's lateralization. For example, a patient with a
clearly lateralized early T-category tumor and a clinically N0 neck is
assigned a 1-2\% risk for occult disease in contralateral LNL II. Under
the same scenario, but \emph{with} mid-sagittal extension, the risk
jumps to almost 7\%.

@@

@@

\caption{}\label{T_f8f73}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

>>>>

\caption{}\label{T_c3b84}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Technically, the introduction of this additional random variable doubles
the space of the hidden states and therefore quadruples the size of the
transition matrix \(\mathbf{A}\). But since only the transition matrix
in case of midline extension depends on the lateralization of the tumor,
we implement it differently and thereby make the computation slightly
more efficient.

First, the overall probabilities to find a patient with a clearly
lateralized tumor or one that extends over the mid-sagittal line after
\(t\) time-steps are respectively given by

@@

Using this, it is straightforward to write down the joint probability of
ipsi- and contralalteral states in case of a tumor that does \emph{not}
extend over the mid-sagittal line similar to the marginalization in
equation~\ref{eq-bilateral-marginal}:

\begin{equation}\phantomsection\label{eq-joint-noext}{
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{False} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot (1 - p_\epsilon)^t \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t, \epsilon=\texttt{False} \right)
}\end{equation}

where for the contralateral evolution we used the transition matrix that
depends on the base spread parameters
\(b_v^{\text{c},\epsilon=\texttt{False}}\).

The case when midline extension is eventually present is more
complicated: We already marginalize over the exact time-step when the
tumor grows over the mid-sagittal line. But whenever that happens, we
also need to change the contralateral transition matrix to use the
increased spread rates from the tumor to the contralateral LNLs, given
by the linear mix in equation~\ref{eq-mixing}. We can achieve the
correct marginalization by iteratively building the joint
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\)
from \(t=0\), where we know that all LNLs are healthy and the tumor is
lateralized:

\[
P \left( \mathbf{X}^\text{c} = \boldsymbol{\xi}_0, \epsilon=\texttt{False} \mid t=0 \right) = 1
\]

Then we consider an arbitrary later time-step \(t=\tau+1\). There are
two possible scenarios we need to marginalize over:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is \(p_\epsilon\) and we use this to weight
  the contralateral state distribution that was so far evolved without
  increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, for
  this case we simply add the same distribution we are computing, but
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

\[
P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau + 1 \right) = \big[ p_\epsilon P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{False} \mid \tau \right) + P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right) \big]^\top \cdot \mathbf{A}^{\text{c},\epsilon=\texttt{True}}
\]

Using this, we can write the joint of ipsi- and contralateral
involvement also for the case of mid-sagittal extension:

\begin{equation}\phantomsection\label{eq-joint-ext}{
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)
}\end{equation}

Subsequently, equation~\ref{eq-joint-noext} and
equation~\ref{eq-joint-ext} can be used to compute the likelihoods of
patients with the respective tumor lateralization. If, for some reason,
it is not known whether the tumor crossed the midline or not, we can
simply add the two distributions to marginalize over midline extension
as well.

@@

\caption{}\label{T_6c7f4}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
computed and observed prevalences for scenarios that illustrate the
model's capability of accounting for the correlation between ipsi- and
contralateral involvement. We show two scenarios: One where the
ipsilateral neck shows no involvement (at least LNLs I to V are healthy,
LNL VII was ignored because data on it is missing for some patients) in
blue and one where at least LNL III was involved in orange. These two
scenarios are plotted for all combinations of T-category (early in top
row, advanced in bottom row) and tumor lateralization (lateralized in
left column, extending over mid-sagittal line in right column).}

@@

In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
ability to capture the correlation between ipsi- and contralateral
involvement. It shows that the prevalence of metastases in the two sides
of the neck is correlated via the time of diagnosis, despite the model
not having any direct connections between the two side. However, there
are some discrepancies in the model's prediction: It cannot quite
capture the correlation of ipsi- and contralateral involvement that is
seen in the data. This is notable for example in the bottom right
subplot where the prevalence is overestimated for the scenario of an
(almost) healthy ipsilateral neck and underestimated when at least LNL
III shows metastases. One reason for this may be the width of the
time-prior: If it was wider, the posterior over the diagnosis time given
the ipsilateral involvement could more flexibly shift to reflect a more
severe contralateral involvement.

@@

\includegraphics{figures/fig-model-risks.png}

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult contralateral LNL II (left), III (middle), and IV (right)
invovlement, shown for some combinations of T-category, tumor
lateralization, and ipsilateral clinical involvement. Unless otherwise
mentioned in the legend, the entire contralateral side was assumed to be
clinically node negative.}

@@

Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
in the contralateral LNLs II, III, and IV for interesting combinations
of these three risk factors. This risk is computed \emph{given} a CT
diagnosis assessing the per-LNL clinical involvement for which we
assumed a sensitivity of 81\% and a specificity of 76\%
\citep{de_bondt_detection_2007}.

\subsubsection{Contralateral LNL II}\label{contralateral-lnl-ii}

The variable impacting the prediction for contralateral level II
involvement in our model is the tumor's lateralization. For example, a
patient with a clearly lateralized early T-category tumor and a
clinically N0 neck is assigned a 1-2\% risk for occult disease in
contralateral LNL II. Under the same scenario, but \emph{with}
mid-sagittal extension, the risk jumps to almost 7\%.

@@

\subsubsection{Contralalteral LNL III}\label{contralalteral-lnl-iii}

As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
for occult disease in the contralateral LNL III only ever crosses a 5\%
threshold if the upstream level II is involved too, resulting in a risk
of 5\% to 6\%. As long as the contralateral neck appears clinically node
negative, however, even an advanced T-category and midline extension do
not yield a risk higher than 2\%.

\subsubsection{Contralateral LNL IV}\label{contralateral-lnl-iv}

Our model's prediction for the contralateral LNL IV show that even in
the most extreme case, with an advanced T-category tumor extending over
the mid-sagittal line as well as metastases in every ipsilateral LNL and
every contralateral upstream level, the risk does not exceed 5\%

@@

\caption{}\label{T_37f78}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\
